Reasoning about the rigor of perl programs: \\ \large gaps, limits to understanding, and means to ameliorate

### 2024-06-23 sciperl/paper.tex
finally done

I will lay out what is currently available, how that fits into a framework for reasoning about programs
in a rigorous fashion, and identify places where our ability to reason could be improved via tooling.
Prototypes of said tools will also be demonstrated.

@@

Nevertheless, this is a problem which others have attempted to solve and done so with some success.
Haskell, ML and Swift all have such type inference systems with the caveat of exponential worst-case execution times\cite{clippert}.
This is because the problem is in fact the 3SAT problem\footnote{also known as the boolean satisfiability problem} which is NP-hard\cite{mltype}.

@@

Some users of other programming languages believe themselves to have obviated the entire class of errors due to improper inputs via strong typing systems.
This of course ignores the reality that poor choice of type (such as an "Any" type, or unsigned when you need signs, etc) immediately voids any such guarantees.
Boundary testing also remains necessary.
However the argument is not entirely without merit, and there are various runtime constraint systems such as \texttt{Type::Tiny} attesting to this\footnote{Along with other efforts, such as those promoted by the Perl Types committee: https://perlcommunity.org/types/}.

Typescript\footnote{a ECMAScript (aka Javascript) code generator} is probably the best example of a dynamic language figuring out how to give you such compile-time guarantees.
Supposing everything else it interacts with is also generated by Typescript\footnote{Rarely ever the case in Javascript projects}, a good degree of confidence can be had.
It's also likely that much of the code therein, particularly the types themselves, can be repurposed towards doing the same with perl.

@@

Since not all subroutines are \textit{Klingon Programming}\footnote{where it is better to \textit{die()} than \textit{return()} in failure.  Also known as "Control-Flow by Exception".  Such subroutines usually behave poorly in threaded contexts.},
such failures tend to go unnoticed when inured in this mental model.
This is why a great many Perl::Critic policies exist to catch precisely such problems caused by external factors largely outside of a subroutine's control.
For example, I have observed policies enforcing code along these lines:
\begin{lstlisting}
# Failing to have a 'do' clause after block eval without
# assignment generally means something is failing silently
eval {...;1;} or do {...}

@@

The sources of these faults outside of a subroutine's control are generally syscalls or FFI.
As such we will need some means of discerning whether or not the usage in a test behaves well under these scenarios:
\begin{itemize}
\item call emits (or is subjected to) a signal (e.g. SIGUSR1, SIGHUP, SIGKILL) either directly or in a child
\item call emits an exception
\item call returns a failure code
\item call appears successful, but nonetheless failed
\end{itemize}
Ideally, this would be done in a semi-automated fashion where one could write a test exercising the expected operational modes, and then re-execute subjected to permutations of these failures.
The approach used by \texttt{Test::MockFile}\cite{mock} is likely one which can be leveraged to inject syscall faults, and everything else can be handled with a universal mock (which would wrap whatever existing mock or sub is called.).

@@

Thankfully there is a solution which uses XS to accomplish this: \texttt{Devel::Probe}\footnote{https://metacpan.org/pod/Devel::Probe}.

@@

This is also another place where AI techniques could be of great utility.
While they are necessarily awful at writing \textit{correct} code \cite{copilot}, they're pretty great at classifying code.
As such we could likely build these lists of "Modules that do largely the same things on CPAN" in a mostly automated fashion.
Currently, the best we have is the "SEE ALSO" section of a module's POD, which is far from comprehensive.

>>>>

I will lay out what is currently available, how that fits into a framework for reasoning about programs
in a rigorous fashion, and identify places where our ability to reason could be improved via tooling.

@@

Nevertheless, this is a problem which others have attempted to solve and done so with some success.
Haskell, ML and Swift all have such type inference systems with the caveat of exponential worst-case compile times\cite{clippert}.
This is because the problem is in fact the 3SAT problem\footnote{also known as the boolean satisfiability problem} which is NP-hard\cite{mltype}.

@@

Some users of other programming languages believe themselves to have obviated the entire class of errors due to improper inputs via strong typing systems.
This of course ignores the reality that poor choice of type (such as an "Any" type, or unsigned when you need signs, etc) immediately voids any such guarantees.
However the argument is not entirely without merit, and there are various runtime constraint systems such as \texttt{Type::Tiny} attesting to this\footnote{Along with other efforts, such as those promoted by the Perl Types committee: https://perlcommunity.org/types/}.

Bolting on strict types to a dynamic language is not a subject to approach lightly, as many features like late binding and reflection become far more difficult.
We should learn from existing successful attempts to do so.

Typescript\footnote{a ECMAScript (aka Javascript) code generator} is the most successful example of a dynamic language figuring out how to bolt-on static types.
Supposing everything else it interacts with is also generated by Typescript\footnote{Rarely ever the case in Javascript projects}, a good degree of confidence can be had.

@@

Since not all subroutines are \textit{Klingon Programming}\footnote{where it is better to \textit{die()} than \textit{return()} in failure.  Also known as "Control-Flow by Exception".  Such subroutines usually behave poorly in threaded contexts.},
failures-by-value tend to go unnoticed when inured in this mental model.
This is why a great many Perl::Critic policies exist to catch precisely such problems caused by external factors largely outside of a subroutine's control.
For example, I have observed policies enforcing code along these lines:
\begin{lstlisting}
# Failing to have a 'do' clause after block eval without
# assignment generally means something is failing silently
eval {...;1;} or do {...}

@@

The sources of these faults outside of a subroutine's control are generally from syscalls or FFI\footnote{Foreign function interface}.
As such we will need some means of discerning whether or not the usage in a test behaves well under these scenarios:
\begin{itemize}
\item call emits (or is subjected to) a signal (e.g. SIGUSR1, SIGHUP, SIGKILL) either directly or in a child
\item call emits an exception
\item call returns a failure code
\item call appears successful, but nonetheless failed
\end{itemize}
Ideally, this would be done in a semi-automated fashion where one could write a test exercising the expected operational modes, and then re-execute subjected to permutations of these failures.
The approach used by \texttt{Test::MockFile}\cite{mock} is likely one which can be leveraged to inject syscall faults, and everything else can be handled with a universal mock (which would wrap whatever existing mock or sub is called.).

@@

Thankfully there is a solution which uses XS to accomplish this: \texttt{Devel::Probe}\footnote{https://metacpan.org/pod/Devel::Probe}.
There is also a more fleshed-out and generic framework called \texttt{Devel::Events}\footnote{https://metacpan.org/pod/Devel::Events}.

@@

This is also another place where AI techniques could be of great utility.
While they are necessarily inept at writing \textit{correct} code \cite{copilot}, they're pretty great at classifying code.
As such we could likely build these lists of "Modules that do largely the same things on CPAN" in a mostly automated fashion.
Currently, the best we have is the "SEE ALSO" section of a module's POD, which is far from comprehensive.
